A family of accessibility measures derived from spatial interaction principles

### 2025-02-09 paper/manuscript.qmd
Editing while proof-reading

---
title: "Accessibility, a type of spatial interaction model" #"A family of accessibility measures (bringing balance back?)"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(flextable)
library(knitr)
library(glue)
library(here)
# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
# library(ggspatial) # Spatial Data Framework for ggplot2
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(here)
```

@@

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs--> 
Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more. In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997]. However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020]. With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling]. However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geursAccessibilityEvaluationLanduse2004; @ferreiraReenactingMobilityAccessibility2020].

<!-- Considering population-level methods in transportation and city planning research, accessibility measures have been used to tackle a wide variety of research questions. They have taken on numerous definitions and analytical expressions; from simpler cumulative opportunities measures, to measures considering more complex travel behaviour and trip weighting characteristics, measures of _potential of opportunities for interaction_ are varied. Summarizing these definitions across studies have been challenging; a testament to these unsettled classifications are present in influential accessibility review literature [@handyMeasuringAccessibilityExploration1997; @geursAccessibilityEvaluationLanduse2004; @wuUnifyingAccess2020; @millerAccessibilityMeasurementApplication2018]. -->

<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs. We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.   -->
In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction roots. It proposes to do so by mapping accessibility onto Wilson's family of spatial interactions models [@wilson1971]. In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its spatial interaction history. To this end, we begin by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to its split into accessibility research from @hansen1959 to today. After this, we derive different types of accessibility measures, following the family of "spatial interaction models" (SIM) namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we will discuss the use of each measure and broadly the implications of ignoring the role of constrains in accessibility analysis.

# Background

## The inquiry into spatial interaction traces to Newton's Universal Gravitation
<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
The patterns of people's movement in space have been a subject of scientific inquiry for centuries. From as far back as Henry C. Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space. It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality only means that as the product of the masses increases, so does the force. Likewise, inverse proportionality only means that as the distance increases, the force decreases. To quantify the magnitude of the force, an empirical factor, say $G$, is required to convert the proportionality into an equality, which ensures that the force $F$ matches empirical observations of the force of attraction between masses. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is a proportionality constant obtained empirically:

@@

Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that Hutton's experiments of 1778 validated [@hutton_xxxiii_1778, p. 783]. Still, it took over a century from the publication of _Principia_ to refine the estimate of this constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].

## Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)

<!-- SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. These researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy. -->
Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts. In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination. These propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav is missing. Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was expressed as follows:
$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

\noindent Where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

Zipf, on the other hand, wrote his hypothesis in mathematical form as:
$$
C^2 = \frac{P_1\cdot P_2}{D}
$${#eq-zipf}

\noindent Where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.

After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948]. We suspect that this may relate to academic backgrounds; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists. Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
$$
F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
$${#eq-stewart-force}

@@

- $F$ is the _demographic force_
- $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
- $d_{12}^2$ is the distance between $1$ and $2$
- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]

In addition to demographic force, Stewart defined a measure of "population potential" of $2$ with respect to $1$ as follows:
$$
V_1 = G\frac{M_2}{d}
$${#eq-stewart-population-potential}

For a system with more than 2 population bodies, Stewart formulated the population potential at $i$ as:
$$
V_i = \int\frac{D}{r} ds
$${#eq-stewart-population-potential-integral}

\noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. In more contemporary terms, this can be rewritten as:
$$
V_i = \sum_j \frac{M_j}{d_{ij}}
$${#eq-stewart-population-potential-sum}

\noindent which astute readers will identify as our modern definition of accessibility.

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p. 34. Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism. For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.

In summary: from the 1850s to the 1940s, researchers attempted to theoretically and empirically characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$. Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.

## Hansen's gravity-based accessibility to today
<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up. Is now cited as the father of accessibility. However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day. The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction... a generalization of the population-over-distance relationship or _population potential_ concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73]. As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
$$
S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
$${#eq-accessibility}

Alternatively, a more general form of this expression to account for different impedance functions:
$$
S_{i} = \sum_j O_j \cdot f(d_{ij})
$${#eq-accessibility-general}

$S_{i}$ in @eq-accessibility is a measure of the accessibility at $i$, whereas $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. Today, Hansen is is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance. Furthermore, those reading @hansen1959 should remember that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$ to 1, with a note that"$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

>>>>

---
title: "A family of accessibility measures derived from spatial interaction principles"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(bibliometrix) # Comprehensive Science Mapping Analysis
library(dplyr) # A Grammar of Data Manipulation
library(flextable) # Functions for Tabular Reporting
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(glue) # Interpreted String Literals
library(gt) # Easily Create Presentation-Ready Display Tables
library(here) # A Simpler Way to Find Your Files
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
# library(ggspatial) # Spatial Data Framework for ggplot2
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(here) # A Simpler Way to Find Your Files
```

@@

<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy. This has had problematic consequences. Conceptually, accessibility is a more complete and comprehensive evolution of mobility. As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->

Historically, the focus of transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement. This has had problematic consequences: with the car seen as the ultimate mobility tool, this approach has led to the emergence and continued dominance of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013]. Decades of mobility-based planning (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often have had limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010]. In response to this situation, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020] using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]. Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape. In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959]. Compared to other measures of performance used in transportation that benchmark movement (e.g., VKT, PKT, etc.), accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].

As a result, scholarly research on accessibility has experienced a remarkable boom, and has grown to include employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application. In other words, accessibility analysis is used today to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020]. However, despite its growth in popularity in scholarly works, challenges remain to more widespread adoption of accessibility in planning practice. Several barriers to bridging the scholarly-practice accessibility gap have been identified. For example, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023. Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.

Adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023]. The menu of accessibility measures has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024]. How is a practitioner to choose among this myriad options? What differences in accessibility scores should matter, and how should they be communicated? [see @vanweeAccessible2016; p. 14].

In this respect, we welcome @wuUnifyingAccess2020's contribution for the way it provides a unifying framework to think about accessibility measures. By considering the key features of accessibility, namely travel-cost and the distribution of opportunities, these authors demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula. One needs only to judiciously change the way travel-cost and opportunities are formulated, to derive almost any known accessibility measure.

With their research, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators. Still, we feel that there is room to further clarify another relevant aspect. @wuUnifyingAccess2020's treatment of competition deserves further scrutiny, since the unifying framework does not appear to accommodate many (or any) of the hugely popular floating catchment area methods. In recent research, however, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn an accessibility measure of the following type (the most general in Wu and Levinson's framework) into a competitive measure of accessibility _à la_ 2-Step Floating Catchment Area approach:

$$
S_i \propto \sum_j g(O_i)f(c_{ij})
$$

In this paper we contend that accessibility research needs to reconnect with its roots in spatial interaction modelling. It is by looking to the past that we believe accessibility analysis can take a new step to the future. In particular, we argue that an important aspect of spatial interaction modelling, namely constraining the results to match empirical observations, was never properly reincorporated into accessibility analysis after these two streams of research began to diverge after the work of @hansen1959 and @wilson1971. As Wilson demonstrated, applying empirical constraints led to various spatial interaction models, and in this work we aim to show that the same can be done for accessibility measures. On the contrary, the lack of constraints in accessibility analysis seems to have contributed to some of the interpretability issues that plague it, particularly the fuzziness of insights beyond making statements of proportion such as higher-than/lower-than.

With this background in mind, the focus of the present paper is on the empirical constraints and the attendant proportionality constant missing from accessibility measures. We propose to map accessibility onto Wilson's family of spatial interactions models. To achieve this, we begin by tracing the development of accessibility from its origins in spatial interaction, beginning with @ravensteinLawsMigration1889 to the accessibility research of @hansen1959. We also provide evidence of a marked divergence of accessibility and spatial interaction modelling research after the work of @wilson1971. After this, we hark back to Wilson's spatial interaction model, and use it to derive a family of accessibility measures based on different types of constraints. We illustrate various members of this family with a simple numerical example and a real world data set. We then conclude by discussing the uses of these measure and their interpretation.

<!--
We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values, as well as converting accessibility indicators from ordinal variables into interval/ration ones. To address this matter, we need not reinvent the wheel. Stewart left the determination of $G$ for the future, possibly expecting it to be some sort of universal constant. However, the solution to this problem was found by @wilson1971, who discovered that there was not a single $G$, but rather various balancing factors that depend on 1) empirical observations of population and/or opportunities; and 2) whether (and how) they were used to constrain the analysis. This development is outlined next.
-->

# Newtonian's roots of human spatial interaction research

<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->

The patterns of people's movement in space have been a subject of scientific inquiry for almost a century and a half. From as far back as Henry C. Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] a concern can be observed with the scientific study of human spatial interaction. It was in this work where Carey stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp. 37-38]. This statement shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation, first posited in 1687's _Principia Mathematica_ and expressed as in @eq-phys-grav-prop.

@@

To be certain, the expression above, a proportionality, is one of the most famous in all of science. In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. Direct proportionality means that as the product of the masses increases, so does the force. Likewise, inverse proportionality means that as the distance increases, the force decreases. @eq-phys-grav-prop, however, does not quantify the magnitude of the force, for to do so an empirical constant is required to convert the proportionality into an equality, thus ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses. In other words, @eq-phys-grav-prop needs to be _constrained_ using empirical data. Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically callibrated proportionality constant:

@@

Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p. 783]. Still, it took over a century from the publication of _Principia_ to refine the estimate of the proportionality constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].

# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)

<!-- 
SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s. A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance. This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-->

Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts. In the late 1880s Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations. As with Carey, these propositions are similar to Newton's gravitational laws. Over time, other researchers discovered similar relationships. <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations. Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways [@zipfHypothesisCaseRailway1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and information [@zipfDeterminantsCirculationInformation1946]. The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.

A common feature of these early investigations of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav was never considered. Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was presented in the following form:

$$
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
$${#eq-reilly}

\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

Zipf, on the other hand, wrote his hypothesis in mathematical form as:

$$
C^2 = \frac{P_1\cdot P_2}{D_{12}}
$${#eq-zipf}

\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.

After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948]. We suspect that this relates to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist. Besides awareness of preceding research (Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:

$$
F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
$${#eq-stewart-force}

@@

- $F$ is the _demographic force_
- $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
- $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
- $d_{12}^2$ is the distance between $1$ and $2$
- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]

In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
$$
V_1 = G\frac{M_2}{d_{12}}
$${#eq-stewart-population-potential}

For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
$$
V_i = \int\frac{D}{r} ds
$${#eq-stewart-population-potential-integral}

\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$. In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
$$
V_i = \sum_j \frac{M_j}{d_{ij}}
$${#eq-stewart-population-potential-sum}

Alerted readers will notice that @eq-stewart-population-potential-sum is formally equivalent to our modern definition of accessibility.

Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic. It had issues with inconsistent mathematical notation. <!--For example, in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant p. 34 and then later as _demographic energy_ on p. 53.--> More seriously, though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism. For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" [p. 35]. Stewart's ideas about "social physics" soon fell out of favour among social scientists, but not before influencing the nascent field of accessibility research, as detailed next.

# Hansen's gravity-based accessibility to today

<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up. Is now cited as the father of accessibility. However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day. The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it. The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient. In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed. -->

From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G. Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959]. In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction... a generalization of the population-over-distance relationship or _population potential_ concept developed by @stewartDemographicGravitationEvidence1948" (p. 73). As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense. @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind. @hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns . In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:

$$
S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
$${#eq-accessibility}

A modern re-writing of the above accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:

$$
S_{i} = \sum_j O_j \cdot f(d_{ij})
$${#eq-accessibility-general}

$S_{i}$ in @eq-accessibility is a measure of the accessibility of site $i$. This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g.,  distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost). Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance. At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determine it. In this way, $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020]. Alas, without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated. Hansen-style accessibility, therefore, is technically an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].

# Wilson's family of spatial interaction models
